---
title: "Himalayan Climbing Expeditions"
author: "James P. Hare"
date: "9/23/2020"
output: github_document
---

https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md

https://www.alexcookson.com/post/how-dangerous-is-climbing-mount-everest/


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

theme_set(theme_minimal())

members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv')
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
peaks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/peaks.csv')
```

"What is the composition of expeditions? For example, how big do they tend to be and what proportion consist of hired staff?

```{r}
everest_expeditions <- expeditions %>% 
        filter(peak_id == "EVER") %>% 
        mutate(size_group = members + hired_staff) %>% 
        group_by(year) %>% 
        summarise(avg_staff = mean (hired_staff),
                  avg_members = mean(members),
                  avg_size = mean(size_group)
                  ) %>% 
        mutate(prop_staff = avg_staff / avg_size)
```

```{r}
everest_expeditions %>% 
        ggplot(aes(year, avg_size)) +
        geom_col()
```

```{r}
everest_expeditions %>% 
        ggplot(aes(year, prop_staff)) +
        geom_col()
```

"What injuries are associated with climbing Everest?"

```{r}
members %>% 
        filter(peak_id == "EVER",
               injured == TRUE) %>%
        group_by(injury_type) %>% 
        count() %>% 
        ggplot(aes(x = reorder(injury_type, n), y = n)) +
        geom_col() +
        coord_flip()
```

"Are there characteristics of climbers associated with higher or lower death rates? For example, are Sherpas – presumably well-acclimated to high altitudes – less likely to suffer from AMS?"

Let's tackle the example question first. First let's take a look at how Sherpas die on Everest. 

```{r}
# Hired staff here seems to be a rough stand-in for what is meant by Sherpa. 
# I'm not sure what the one Belgian is doing on this list, but they're not among
# the deaths, so I think we're pretty safe using this definition.

members %>% 
        filter(hired == TRUE) %>% 
        group_by(citizenship) %>% 
        count() %>% 
        arrange(desc(n))
```


```{r}
sherpa_deaths <- members %>% 
        filter(peak_id =="EVER",
               hired == TRUE,
               died == TRUE) %>%
        group_by(death_cause) %>% 
        count() %>% 
        arrange(desc(n))

sherpa_deaths
```

Now let's look at how non-Sherpas die on Everest.

```{r}
non_sherpa_deaths <- members %>% 
        # So non-Sherpa here means everyone except Nepalese citizens working as hired staff
        filter(peak_id =="EVER",
               died == TRUE,
               hired == FALSE
        ) %>% 
        group_by(death_cause) %>% 
        count() %>% 
        arrange(desc(n))

non_sherpa_deaths
```

So among deaths on Everest, Sherpas, who are required to spend much more time in dangerous zones, are much more likely to die in avalanches while non-Sherpas, who are presumably less well acclimated to the altitude, face a greater risk of AMS.

Out of 120 Sherpa deaths on Everest, 56 (46.7%) died in avalanches while 9 (7.5%) died from AMS.

Out of 186 non-Sherpa deaths, 27 (14.5%) died from AMS. The most common cause of death for non-Sherpas was falling. 54 of these climbers (29%) died from falls.

```{r}
members %>% 
        group_by(hired) %>% 
        count()
```

However, if we look at it based on total number of climbers, then the risk of death from altitude sickness is actually marginally worse for Sherpas than non-Sherpas, meaning that Sherpas are facing a comparable risk from the altitude as well as much greater risks from avalanches and other dangers.
