---
title: "Plants in Danger"
author: "James Hare"
date: "8/18/2020"
output: github_document
---

## Extinct Plants

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

#First load needed libraries and this week's data

library(tidyverse)

plants <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/plants.csv')
actions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/actions.csv')
threats <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/threats.csv')
```


```{r}

#Then wrangle the data

year_levels <- c("Before 1900", "1900-1919", "1920-1939", "1940-1959", 
                  "1960-1979", "1980-1999","2000-2020")

plants <- plants %>% 
        mutate(year_last_seen = factor(year_last_seen, levels = year_levels))

threats <- threats %>% 
        mutate(year_last_seen = factor(year_last_seen, levels = year_levels)) %>%
        filter(threatened == 1)
        
actions <- actions %>% 
        mutate(year_last_seen = factor(year_last_seen, levels = year_levels)) %>%
        filter(action_taken == 1)
```

Let's start by visualizing the data in various ways to see if we can find anything interesting.


```{r}
plants %>% 
        ggplot(aes(year_last_seen)) +
        geom_bar() +
        coord_flip()
```


```{r}

plants %>% 
        ggplot(aes(year_last_seen, fill = continent)) +
        geom_bar(position = "dodge") +
        coord_flip()

```

```{r}
plants %>%
        filter(!(is.na(year_last_seen))) %>% 
        group_by(year_last_seen, continent) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n, group = continent)) +
        geom_line(aes(color = continent))

```

```{r}

plants %>% 
        ggplot(aes(year_last_seen)) +
        geom_bar() +
        facet_wrap(~ continent) +
        coord_flip()

```

```{r}

threats %>% 
        ggplot(aes(year_last_seen, fill = threat_type)) +
        geom_bar(position = "dodge") +
        coord_flip()

```
```{r}
threats %>%
        filter(!(is.na(year_last_seen))) %>% 
        group_by(year_last_seen, threat_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n, group = threat_type)) +
        geom_line(aes(color = threat_type))

```

```{r}

actions %>% 
        ggplot(aes(year_last_seen, fill = action_type)) +
        geom_bar(position = "dodge") +
        coord_flip()

```
```{r}
actions %>%
        filter(!(is.na(year_last_seen))) %>% 
        group_by(year_last_seen, action_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n, group = action_type)) +
        geom_line(aes(color = action_type))
```

```{r}
threats %>%
        filter(threatened == 1) %>% 
        group_by(continent, threat_type) %>% 
        count() %>%
        ggplot(aes(threat_type, n)) +
        geom_col() +
        facet_wrap(~ continent) +
        coord_flip()
```

```{r}
#A lot of plant extinction has happened in Africa. How much of this is Madagascar (which accounts for 19.6% of extinct plant species)

plants %>% 
        filter(continent == "Africa") %>% 
        count()

plants %>% filter(country == "Madagascar") %>% 
        count()

#So 218 out of 500 extinct plant species are from Africa and 98 of these are from Madagascar

#Let's try narrowing our focus to Africa and Madagascar to see if we can say anything interesting about the data

threats %>% 
        filter(continent == "Africa",
               threatened == 1) %>%
        group_by(year_last_seen, threat_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col(aes(fill = threat_type), position = "dodge")

```

```{r}
threats %>% 
        filter(country == "Madagascar",
               threatened == 1) %>%
        group_by(year_last_seen, threat_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col(aes(fill = threat_type), position = "dodge")
```

```{r}
threats %>% 
        filter(continent == "Africa") %>% 
        ggplot(aes(threat_type)) +
        geom_bar() +
        coord_flip()
```

```{r}
threats %>% 
        filter(country == "Madagascar") %>% 
        ggplot(aes(threat_type)) +
        geom_bar() +
        coord_flip()
```

```{r}
plants  %>%  
        filter(continent == "Africa") %>%
        group_by(year_last_seen) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col()
```

```{r}
plants  %>%  
        filter(country == "Madagascar") %>%
        group_by(year_last_seen) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col()
```

```{r}
actions %>% 
        filter(continent == "Africa",
               action_taken == 1) %>%
        group_by(year_last_seen, action_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col(aes(fill = action_type), position = "dodge")
```

```{r}
actions %>% 
        filter(country == "Madagascar",
               action_taken == 1) %>%
        group_by(year_last_seen, action_type) %>% 
        count() %>% 
        ggplot(aes(year_last_seen, n)) +
        geom_col(aes(fill = action_type), position = "dodge")
```

I also tried mapping some of the data, but sorting the data by country isn't that interesting and perhaps deceptive; although, it does show that islands are the site of many plant extinctions.

So the stories I see here that are worth telling is how much plant extinction has taken place in Africa and Madagascar in particular, how much of this took place during the mid-twentieth century, and how much of it is due to "Agriculture and Aquaculture" and "Biological Resource Use" (and to a slightly lesser degree "Natural System Modifications"). It's also interesting how much loss occurred in South America prior to 1900. So now let's figure out how to tell this story visually.

```{r}
#Define shared plot aesthetic
theme_set(theme_minimal())

centuries <- c("Before 1900", "After 1900")

mad_plants <- plants %>% 
        filter(!(is.na(year_last_seen))) %>%
        mutate(pre1900 = if_else(year_last_seen == "Before 1900", 
                                 "Before 1900", "After 1900"),
               pre1900 = factor(pre1900, levels = centuries)) %>% 
        filter(country == "Madagascar")

plants %>% 
        filter(!(is.na(year_last_seen))) %>%
        mutate(pre1900 = if_else(year_last_seen == "Before 1900", 
                                 "Before 1900", "After 1900"),
               pre1900 = factor(pre1900, levels = centuries)
                       ) %>% 
        ggplot(aes(reorder(continent, desc(continent)))) +
        geom_bar() +
        geom_bar(data = mad_plants, aes(fill = "Madagascar")) +
        facet_wrap(~ pre1900) +
        coord_flip() +
        labs(x = NULL,
             y = NULL,
             title = "Most Plant Extinction Has Occurred Since 1900",
             subtitle = "Africa, especially Madagascar, has had a disproportionate share" 
             ) +
        theme(legend.position = "bottom",
              legend.title = element_blank())

mad_plants  %>%  
        group_by(year_last_seen) %>% 
        count() %>%  
        ggplot(aes(year_last_seen, n)) +
        geom_col() +
        labs(x = "Year Last Seen",
             y = NULL,
             title = "Madagascar Lost the Most Plant Species During the Mid-Twentieth Century"
             ) +
        theme(legend.position = "none")

threats %>% 
        filter(country == "Madagascar") %>% 
        ggplot(aes(reorder(threat_type, desc(threat_type)))) +
        geom_bar() +
        coord_flip() +
        labs(x = NULL,
             y = NULL,
             title = "Habitat Destruction Has Caused the\nBulk of Madagascar's Plant Extinction",
             caption = "Source: International Union for Conservation of Nature" 
             ) +
        theme(legend.position = "none")
```
